---
title: "MDS for Linguists"
output:
  md_document:
    variant: markdown_github
    toc: TRUE
    toc_depth: 2
---

  
## MDS for Linguists

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("/home/jtimm/pCloudDrive/GitHub/git-projects/render_toc.r")
```

**An R-based guide for linguistic typologists** interested in applying [NOMINATE](https://voteview.com/about) multidimensional scaling (MDS) techniques to linguistic data as presented in [Croft](http://www.unm.edu/~wcroft/) and [Poole](https://polisci.ucsd.edu/about-our-people/faculty/faculty-directory/emeriti-faculty/poole-profile.html), "Inferring universals from grammatical variation: multidimensional scaling for typological analysis" (*Theoretical Linguistics* 34.1-37, 2008)." [[Abstract]](https://www.degruyter.com/view/j/thli.2008.34.issue-1/thli.2008.001/thli.2008.001.xml)


This guide provides a brief summary of an R-based workflow for model implementation and the visualization of model results within the `ggplot` data visualization framework.  A cross-linguistic data set of indefinite pronouns from Haspelmath (1997) is utilized (and made available) here for demonstration purposes. For more thoughtful discussions regarding theory, scaling procedures & model interpretation, see reference section.


```{r echo=FALSE}
render_toc("/home/jtimm/pCloudDrive/GitHub/git-projects/mds_for_linguists/README.Rmd", 
           toc_header_name = 'MDS for Linguists',
           toc_depth = 2)
```



## Getting started

### Install and load required packages

```{r message=FALSE, warning=FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(# anominate, -- no longer maintained -- 
               wnominate, 
               pscl, 
               ggplot2, 
               knitr, 
               devtools, 
               ggrepel, 
               data.table)
```


```{r eval=FALSE}
devtools::install_github("jaytimm/wnomadds")
library(wnomadds)
```


### Load data

Data set:  A 9 x 140 data frame: Nine indefinite pronominal meanings, using data from 140 pronouns in 40 languages.  Data are made available [here](https://github.com/jaytimm/mds_for_linguists_using_R/blob/master/resources/Indefprn13.txt).


```{r include=FALSE}
local_data <- '/home/jtimm/pCloudDrive/GitHub/git-projects/mds_for_linguists/resources'
```


```{r eval=FALSE}
## File paths will look differently for Windows/Mac
local_data <- '/home/jtimm/Desktop/data/'
```


Load data set:

```{r message=FALSE, warning=FALSE}
setwd(local_data)
raw_data <- read.csv("Indefprn13.txt",
            sep="\t", 
            stringsAsFactors = FALSE)
```



A portion of the data frame is presented below.  Rows contain functions/meanings, and are analagous to legislators in the NOMINATE model.  Columns contain language-specific grammatical forms, and are analagous to roll calls (ie, votes) in the NOMINATE model.  

A value of 1 in the table below means that a given form expresses a particular meaning; a value of 6 means that a given form does not express that particular meaning.  Missing data are specified with the value 9.


```{r message=FALSE, warning=FALSE}
knitr::kable(raw_data[,1:9]) 
```


---


## Using `wnominate` & `pscl`

### Building MDS models

#### Rollcall object

The first step is to transform the original data structure into a `rollcall` object using the `pscl` package.


```{r}
roll_obj <- pscl::rollcall(raw_data [,-1], 
                           yea=1, 
                           nay=6, 
                           missing=9,
                           notInLegis=8,
                           vote.names = colnames(raw_data)[2:ncol(raw_data)], 
                           legis.names = raw_data[,1])
```



#### Ideal points estimation 

Then we fit three models using the `wnominate` function -- one-, two- & three-dimensional solutions.

```{r message=FALSE, warning=FALSE, results = 'hide'}
ideal_points_1D <- wnominate::wnominate (roll_obj, dims = 1, polarity=c(1)) 
ideal_points_2D <- wnominate::wnominate (roll_obj, dims = 2, polarity=c(1,2)) 
ideal_points_3D <- wnominate::wnominate (roll_obj, dims = 3, polarity=c(1,2,3)) 
```


The resulting data structures are each comprised of seven elements:

```{r}
names(ideal_points_1D)
```



#### Model comparison and fitness statistics

Correct classification and fitness statistics for each model are extracted from the `fits` element, and summarized below: 

```{r message=FALSE, warning=FALSE}
list('1D' = ideal_points_1D$fits, 
     '2D' = ideal_points_2D$fits, 
     '3D' = ideal_points_3D$fits)
```



### Visualizing model results


#### A one-dimensional solution 

Extract legislator coordinates (ie, ideal points) from one-dimensional model results.

```{r}
d1 <- cbind(label=rownames(ideal_points_1D$legislators), 
            ideal_points_1D$legislators)
d1 <- d1[order(d1$coord1D),]
```



Plot legislators (ie, grammatical functions) in one-dimensional space by rank.

```{r fig.height=6, fig.width=6}
ggplot()  +
  geom_text(data = d1,
            aes(x=reorder(label, coord1D), 
                y=coord1D, 
                label=label), 
            size=4, 
            color = 'blue') +
  
#  theme_classic() +
   theme_minimal() +
  
  labs(title="1D W-NOMINATE Plot") + 
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank())+ 
  xlab('') + ylab('First Dimension')+
  ylim(-1.1, 1.1)+
  coord_flip()
```




#### A two-dimensional solution 

We first build a simple "base" plot using legislator coordinates from two-dimensional model results.  Per `wnominate` convention, we add a unit circle to specify model constraints.  All subsequent plots are built on this simple base plot. 


```{r fig.height=6, fig.width=6}
base_2D <- ggplot(data = ideal_points_2D$legislators,
       aes(x=coord1D, 
           y=coord2D)) +
  
  geom_point(size= 1.5,
             color = 'blue') +
  
  annotate("path",
           x=cos(seq(0,2*pi,length.out=300)),
           y=sin(seq(0,2*pi,length.out=300)),
           color='gray',
           size = .25) +
  
  xlab('First Dimension') + 
  ylab('Second Dimension') 

base_2D + ggtitle('Two-dimensional base plot')
```



**Add** labels, a title, and change the theme.  

```{r fig.height=6, fig.width=6}
base_2D +

  ggrepel::geom_text_repel(
    data  = ideal_points_2D$legislators,
    aes(label = rownames(ideal_points_2D$legislators)),
    direction = "y",
    hjust = 0, 
    size = 4,
    color = 'blue') +
  
  theme_classic() +
# theme_minimal() +
  
  ggtitle("W-NOMINATE Coordinates") 
```


---


### Cutting lines and roll call polarity via `wnomadds`

I have developed a simple R package, `wnomadds`, that facilitates the plotting of roll call cutting lines and roll call polarities using `ggplot`.  While `wnominate` provides functionality for plotting cutting lines, only plotting in base R is supported.  The `wnm_get_cutlines` function extracts cutting line coordinates from `wnominate` model results, along with coordinates specifying the direction of majority Yea votes for a given roll call (ie, vote polarity). Addtional details about the package are available [here](https://github.com/jaytimm/wnomadds).

```{r message=FALSE, warning=FALSE}
with_cuts <- wnomadds::wnm_get_cutlines(ideal_points_2D, 
                                        rollcall_obj = roll_obj, 
                                        arrow_length = 0.05)
```


A sample of the resulting data frame:

```{r}
head(with_cuts)
```



#### Cutting lines & legislator coordinates

```{r fig.height=6, fig.width=6}
base_2D +

  ggrepel::geom_text_repel(
    data  = ideal_points_2D$legislators,
    aes(label = rownames(ideal_points_2D$legislators)),
    direction = "y",
    hjust = 0, 
    size = 4,
    color = 'blue') +
  
  geom_segment(data = with_cuts, 
               aes(x = x_1, y = y_1, 
                   xend = x_2, yend = y_2),
               size = .25) + #cutting start to end
  
  theme_minimal() +
  labs(title="Cutting lines & W-NOMINATE Coordinates")
```



#### Cutting lines, roll call polarity & legislator coordinates

```{r fig.height=6, fig.width=6}
base_2D +
  
  geom_segment(data=with_cuts, 
               aes(x = x_1, y = y_1, 
                   xend = x_2, yend = y_2),
               size = .25) + #cutting start to end
  
  ##ARROWS -- 
  geom_segment(data=with_cuts, 
               aes(x = x_2, y = y_2, 
                   xend = x_2a, yend = y_2a), 
               #cutting end to opposite arrow
               color = 'red',
               arrow = arrow(length = unit(0.2,"cm"))) +
  
  geom_segment(data=with_cuts, 
               aes(x = x_1, y = y_1, 
                   xend = x_1a, yend = y_1a), 
               #cutting start to opposite arrow
               color = 'red',
               arrow = arrow(length = unit(0.2,"cm")))+ 
  ##END ARROWS.
  
  geom_text(data=with_cuts, 
               aes(x = x_1a, y = y_1a, 
                   label = Bill_Code), 
               size=2.5, 
               nudge_y = 0.03,
               check_overlap = TRUE) +

  theme_minimal() +
  labs(title = "W-NOMINATE Coordinates, cutting lines & roll call polarity")
```




#### Selected cutting lines and legislator coordinates

```{r fig.height=6, fig.width=6}
selected <- c('X01e', 'X01j', 'X01jd', 'X01n')

subset_cuts <- subset(with_cuts, Bill_Code %in% selected)

base_2D +
  
  ggrepel::geom_text_repel(
    data  = ideal_points_2D$legislators,
    aes(label = rownames(ideal_points_2D$legislators)),
    direction = "y",
    hjust = 0, 
    size = 4,
    color = 'blue') +
  
  geom_segment(data=subset_cuts, 
               aes(x = x_1, y = y_1, 
                   xend = x_2, yend = y_2),
               size = .25) + #cutting start to end
  
  ##ARROWS -- 
  geom_segment(data=subset_cuts, 
               aes(x = x_2, y = y_2, 
                   xend = x_2a, yend = y_2a), 
               #cutting end to opposite arrow
               color = 'red',
               arrow = arrow(length = unit(0.2,"cm"))) +
  
  geom_segment(data=subset_cuts, 
               aes(x = x_1, y = y_1, 
                   xend = x_1a, yend = y_1a), 
               #cutting start to opposite arrow
               color = 'red',
               arrow = arrow(length = unit(0.2,"cm")))+ 
  ##END ARROWS.
  
  geom_text(data=subset_cuts, 
               aes(x = x_1a, y = y_1a, 
                   label = Bill_Code), 
               size=2.5, 
               nudge_y = 0.03,
               check_overlap = TRUE) +

  theme_minimal() +
  labs(title = "W-NOMINATE Coordinates & selected cutting lines")
```



### Facet cutting lines by language

```{r fig.height=6}
#Extract language code from language-specific grammatical forms
with_cuts$lang <- gsub('[A-Za-z]', '', with_cuts$Bill_Code)

#Filter cutting line data set to first six language codes.
facet_cuts <- subset(with_cuts, lang %in% c('01', '02', '03', '04', '05', '06'))

base_2D +
  
  geom_segment(data=facet_cuts, 
               aes(x = x_1, y = y_1, 
                   xend = x_2, yend = y_2),
               size = .25) + #cutting start to end
  
  ##ARROWS -- 
  geom_segment(data=facet_cuts, 
               aes(x = x_2, y = y_2, 
                   xend = x_2a, yend = y_2a), 
               #cutting end to opposite arrow
               color = 'red',
               arrow = arrow(length = unit(0.2,"cm"))) +
  
  geom_segment(data=facet_cuts, 
               aes(x = x_1, y = y_1, 
                   xend = x_1a, yend = y_1a), 
               #cutting start to opposite arrow
               color = 'red',
               arrow = arrow(length = unit(0.2,"cm")))+ 
  ##END ARROWS.

  theme_minimal() +
  facet_wrap(~lang) +
  coord_fixed()+
  labs(title = "W-NOMINATE Coordinates & language-specific cutting lines")
```




---


## References

Royce Carroll, Christopher Hare, Jeffrey B. Lewis, James Lo, Keith T. Poole and Howard Rosenthal (2017). Alpha-NOMINATE: Ideal Point Estimator. R package version 0.6. URL http://k7moa.c
om/alphanominate.htm

Croft, W., & Poole, K. T. (2008). Inferring universals from grammatical variation: Multidimensional scaling for typological analysis. *Theoretical linguistics*, 34(1), 1-37.

Haspelmath, M. (1997). *Indefinite pronouns*. Oxford: Clarendon Press.

Poole, K. T. (2005). *Spatial models of parliamentary voting*. Cambridge University Press.

Keith Poole, Jeffrey Lewis, James Lo, Royce Carroll (2011). Scaling Roll Call Votes with wnominate in R. *Journal of Statistical Software*, 42(14), 1-21. URL http://www.jstatsoft.org/v42/i14/.

